Equalizer\\Programming and User Guide

### 2009-09-09 Developer/ProgrammingGuide/paper.tex
Minor PUG updates based on print version
[ ] Will break Win32 build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.6 for Equalizer 0.9\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{Installing Equalizer and running \textsf{eqPly}}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{ A 2D compound}}
\end{wrapfigure}
The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}\small A database compound}}
\end{wrapfigure}
The application has to partition the database so that the rendering units
render only part of the database. Some OpenGL features do not work
correctly (anti-aliasing) or need special attention (transparency).

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A stereo compound}}
\end{wrapfigure}
Eye compounds in Equalizer are configured using the \textsf{eye}
parameter, limiting the child to render the \textsf{[ LEFT ]} or
\textsf{[ RIGHT ]} eye. Each child compound uses an output frame, which
is connected to an input frame on the destination channel. The
destination channel can also be used to render an eye pass, in which case
it renders in the correct stereo buffer and no output frame is needed.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{ A DPlex compound}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A pixel compound}}
\end{wrapfigure}
OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at runtime to adjust the rendering accordingly.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel compound kernel}}
\end{wrapfigure}
Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
  {\caption{\label{fPixelKernels}Example pixel kernels for a
      four-to-one pixel compound}}
\end{figure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/cave.pdf}
  {\caption{\label{fConfig}An example configuration}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and projection parameters}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A canvas using four, non-consecutive segments}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A layout with four views}}
\end{wrapfigure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A display wall using a six-segment canvas with a two-view layout}}
\end{figure}

@@

\paragraph{Load Equalizer}
\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D load-balancing}}
\end{wrapfigure}
While pixel and stereo compounds are naturally load-balanced, 2D and DB
compounds often need load-balancing for optimal rendering
performance. Using a load equalizer is transparent to the application,
and can be used on any application for 2D, and on most applications for
DB load-balancing. Some applications do not support dynamic updates of
the database range, and therefore cannot be used with DB load-balancing.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-segment
      load-balancing for two segments using eight GPU's}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\begin{wrapfigure}{l}{.382\textwidth} 
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-segment
        load-balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPU's. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPU's from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
  {\caption{\label{fMonitorEq}\small Monitoring a projection wall}}
\end{wrapfigure}
A monitor equalizer adjusts the frame zoom of the output frames used to
observe the rendering, depending on the destination channel size. The
output frames are downscaled on the GPU before readback, which results
in optimal performance.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace layering for eqPly and eqServer}}
\end{wrapfigure}
Examples shipped with Equalizer are implemented in their own namespaces,
e.g., \textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing
classes from the \textsf{eq} namespace, with the occasional usage of
functionality from the \textsf{eq::net} and \textsf{eq::base} namespace.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified execution model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

\section{\label{sEqPly}The eqPly polygonal renderer}

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/uml}
  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes}}
\end{figure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. For example, we have observed a speedup of 15\% on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\begin{wrapfigure}{r}{.382\textwidth}
%\begin{figure}[h!t]
  \subfigure[]{\label{fObjectDist1}
    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
  \subfigure[]{\label{fObjectDist2}
    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
  \subfigure[]{\label{fObjectDist3}
    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
  {\caption{\label{fObjectDist}Object distribution using subclassing,
      proxies or multiple inheritance}}
\end{wrapfigure}
Distributed objects are addressed using session-unique identifiers,
because pointers to other objects cannot be distributed directly; they
have no meaning on remote nodes. The session used to register
distributed objects is normally the \textsf{eq::Config}, which is
sub-classed from \textsf{eq::net::Session}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination view of a DB compound using
      demonstrative coloring}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the frame data to ensure
that all channels render a given frame using the same position.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one node process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{figure}[h!t]\center
  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
  {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
    thread synchronization models}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-node frame synchronization}}
\end{wrapfigure}
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/mainloop.pdf}
  {\caption{\label{fMainLoop}Synchronization of frame tasks}}
\end{figure}

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fMono}
    \includegraphics[height=2.5cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=2.5cm]{images/stereo}}\hfil
  \subfigure[]{\label{fTracked}
    \includegraphics[height=2.5cm]{images/tracked}}
  \subfigure[]{\label{fHMD}
    \includegraphics[height=2.5cm]{images/hmd}}%
  {\caption{\label{fImmersive}Monoscopic, Stereoscopic, Tracked and HMD
    frusta}}
\end{figure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}
  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML hierarchy of \textsf{eqPly::View}}}
  \vspace{-5ex}
\end{wrapfigure}
Changing view frustum parameters is typically done for non-fullscreen
application windows and multi-view layouts, where the rendering is not
meant to be viewed in real-world size in an immersive environment. A
typical use case is changing the field-of-view of the rendering.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signaled by the
output frame after readback.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional diagram of the compositor}}
\end{wrapfigure}
On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the pre-integrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics}
  {\caption{\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\label{fNetNode}Communication between two nodes}}
\end{wrapfigure}
\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
  {\caption{\label{fNetObject}Communication between two objects}}
\end{wrapfigure}
Distributed objects provide data synchronization functionality, as
described in \sref{sNetObject}. Additionally, applications can send
command packets by creating custom object packets with a command greater
than \textsf{CMD\_OBJECT\_CUSTOM}.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.0 for Equalizer 0.9+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{Installing Equalizer and Running \textsf{eqPly}}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{ A 2D Compound}}
\end{wrapfigure}
The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}\small A Database Compound}}
\end{wrapfigure}
The application has to partition the database so that the rendering units
render only part of the database. Some OpenGL features do not work
correctly (anti-aliasing) or need special attention (transparency).

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/EYE.pdf}
  {\caption{\label{fStereoCmp}\small A Stereo Compound}}
\end{wrapfigure}
Eye compounds in Equalizer are configured using the \textsf{eye}
parameter, limiting the child to render the \textsf{[ LEFT ]} or
\textsf{[ RIGHT ]} eye. Each child compound uses an output frame, which
is connected to an input frame on the destination channel. The
destination channel can also be used to render an eye pass, in which case
it renders in the correct stereo buffer and no output frame is needed.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{ A DPlex Compound}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A Pixel Compound}}
\end{wrapfigure}
OpenGL functionality influenced by the raster position will not work
correctly with pixel compounds, or needs at least special
attention. Among them are: lines, points, sprites, glDrawPixels,
glBitmap, glPolygonStipple. The application can query the current pixel
parameters at runtime to adjust the rendering accordingly.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
\end{wrapfigure}
Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

The destination channel can also be used as a source channel. Contrary
to the other compound modes, it also has to use an output and
corresponding input frame. During rendering, the frustum is 'squeezed'
to configure the pixel decomposition. The destination channel can
therefore not be rendered in place, like with the other compound modes.
\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/pixelKernels.pdf}
  {\caption{\label{fPixelKernels}Example Pixel Kernels for a
      four-to-one Pixel Compound}}
\end{figure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/cave.pdf}
  {\caption{\label{fConfig}An Example Configuration}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A Canvas using four Segments}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A Layout with four Views}}
\end{wrapfigure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/wallLayout.jpg}
  {\caption{A Display Wall using a six-Segment Canvas with a two-View Layout}}
\end{figure}

@@

\paragraph{Load Equalizer}
\vspace{-2ex}\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
\end{wrapfigure}
While pixel and stereo compounds are naturally load-balanced, 2D and DB
compounds often need load-balancing for optimal rendering
performance. Using a load equalizer is transparent to the application,
and can be used on any application for 2D, and on most applications for
DB load-balancing. Some applications do not support dynamic updates of
the database range, and therefore cannot be used with DB load-balancing.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPU's}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\begin{wrapfigure}{l}{.382\textwidth} 
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPU's. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPU's from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/monitorEq.pdf}
  {\caption{\label{fMonitorEq}Monitoring a Projection Wall}}
\end{wrapfigure}
A monitor equalizer adjusts the frame zoom of the output frames used to
observe the rendering, depending on the destination channel size. The
output frames are downscaled on the GPU before readback, which results
in optimal performance.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
\end{wrapfigure}
Examples shipped with Equalizer are implemented in their own namespaces,
e.g., \textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing
classes from the \textsf{eq} namespace, with the occasional usage of
functionality from the \textsf{eq::net} and \textsf{eq::base} namespace.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\label{fModel}Simplified Execution Model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

\section{\label{sEqPly}The eqPly Polygonal Renderer}

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/uml}
  {\caption{\label{fUml}UML Diagram eqPly and relevant Equalizer Classes}}
\end{figure}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and Asynchronous Execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. For example, we have observed a speedup of 15\% on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

\begin{wrapfigure}{r}{.382\textwidth}
%\begin{figure}[h!t]
  \subfigure[]{\label{fObjectDist1}
    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
  \subfigure[]{\label{fObjectDist2}
    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
  \subfigure[]{\label{fObjectDist3}
    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
      Proxies or Multiple Inheritance}}
\end{wrapfigure}
Distributed objects are addressed using session-unique identifiers,
because pointers to other objects cannot be distributed directly; they
have no meaning on remote nodes. The session used to register
distributed objects is normally the \textsf{eq::Config}, which is
sub-classed from \textsf{eq::net::Session}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination View of a DB Compound using
      Demonstrative Coloring}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the frame data to ensure
that all channels render a given frame using the same position.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one Node Process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{figure}[h!t]\center
  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
  {\caption{\label{fThreadModels}Async, draw\_sync and local\_sync
    Thread Synchronization Models}}
\end{figure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/mainloop.pdf}
  {\caption{\label{fMainLoop}Synchronization of Frame Tasks}}
\end{figure}

@@

\begin{figure}[h!t]
  \subfigure[]{\label{fMono}
    \includegraphics[height=2.5cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=2.5cm]{images/stereo}}\hfil
  \subfigure[]{\label{fTracked}
    \includegraphics[height=2.5cm]{images/tracked}}
  \subfigure[]{\label{fHMD}
    \includegraphics[height=2.5cm]{images/hmd}}%
  {\caption{\label{fImmersive}Monoscopic, Stereoscopic, Tracked and HMD
    Frusta}}
\end{figure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}
  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
  \vspace{-5ex}
\end{wrapfigure}
Changing view frustum parameters is typically done for non-fullscreen
application windows and multi-view layouts, where the rendering is not
meant to be viewed in real-world size in an immersive environment. A
typical use case is changing the field-of-view of the rendering.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
\end{wrapfigure}
The frame synchronization through the frame data allows the input frame
to wait for the pixel data to become ready, which is signaled by the
output frame after readback.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[h!t]\center
  \includegraphics[width=.618\textwidth]{images/compositor.pdf}
  {\caption{\label{fCompositor}Functional Diagram of the Compositor}}
\end{wrapfigure}
On a high level, the compositor combines multiple input frames using 2D
tiling, depth-compositing for polygonal data or sorted, alpha-blended
compositing for semitransparent volumetric data. These operations
composite either directly all images on the GPU, or use a CPU-based
compositor and then transfer the pre-integrated result to the GPU. The
high-level entry points automatically select the best algorithm. The
CPU-based compositor uses, based on availability, OpenMP and the
Paracomp compositing library to accelerate its operation.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics}
  {\caption{\label{fStatistics}Statistics for a two-Node 2D Compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\label{fNetNode}Communication between two Nodes}}
\end{wrapfigure}
\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
  {\caption{\label{fNetObject}Communication between two Objects}}
\end{wrapfigure}
Distributed objects provide data synchronization functionality, as
described in \sref{sNetObject}. Additionally, applications can send
command packets by creating custom object packets with a command greater
than \textsf{CMD\_OBJECT\_CUSTOM}.
